Something about cetacean heart rates and ballistocardiograms

### 2021-09-09 analysis/paper/paper.Rmd
Use distance threshold for jerk peak classification

>>>>

```{r bradycardia-relax, fig.cap="Heartrates observed in the ballistocardiogram exhibit characteristic diving physiology patterns. Bradycardia is greatest at the start of the dive (~4-5 bpm), increasing towards the end (~8-9 bpm)."}
bw10hz <- bw180905_53_10hz %>%
  mutate(
    dive_id = split_dives(dt, depth,
                          surface = 2, min_depth = 10, min_dur = 5 * 60),
    dive_norm = normalize_dives(dive_id)
  )
bw_400hz$dive_norm <- approx(bw10hz$dt, bw10hz$dive_norm, bw_400hz$dt)$y

bpm_data <- bw_400hz %>%
  drop_na(bcg_bpm)

bpm_ts <- RobustLinearReg::theil_sen_regression(bcg_bpm ~ dive_norm, bpm_data)

bpm_pred <- tibble(dive_norm = seq(0, 1, length.out = 100)) %>%
  mutate(bcg_bpm = predict(bpm_ts, newdata = .))

ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
  geom_point() +
  geom_line(data = bpm_pred, color = "blue", size = 1.5) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Dive time (normalized)",
       y = "Heartrate (bpm)") +
  expand_limits(y = 0) +
  theme_classic()
```

The magnitude of bradycardia was strongest at the start of dives and relaxed towards the end. On average, heartrate increased from `r round(predict(bpm_ts, data.frame(dive_norm = 0)), 1)` bpm at the start of dives to `r round(predict(bpm_ts, data.frame(dive_norm = 1)), 1)` bpm by the end (Theil-Sen regression, $p < 10^{-10}$).

```{r jerk-peaks, fig.cap="Major peak selection procedure for identifying heartbeats. A Even after smoothing, the norm of the jerk vector (line) has peaks that correspond to heartbeats (solid points) and artifacts (hollow points). We used a data-driven approach to retaining only major peaks. 75 s data excerpt shown as example. B In height-prominence space, peaks form two clusters corresponding to valid heartbeats (solid) and artifacts (hollow). C To separate the clusters, we calculated each peak's distance from the greatest peak (triangle); retaining all peaks closer to the greatest peak (solid circles) than the valley in the bimodal distance distribution (red dashed line) and rejecting the farther peaks (hollow circles). D The bi-modal distribution of peak distances. We selected the the valley in the distance density (blue line) for the distance threshold (red dashed line). Solid and hollow histogram bars correspond to retained and rejected peaks, respectively."}
# Visualize peak thresholding algorithm
jerk <- bw_400hz$njerk_smooth
jerk[is.na(bw_400hz$region_id)] <- 0
window_s <- 1.8
fs_hz <- 400

# Find peaks
peaks <- pracma::findpeaks(jerk, minpeakdistance = window * fs_hz)
jerk_peaks <- slice(bw_400hz, peaks[, 2]) %>% 
  mutate(
    height = peaks[, 1],
    prominence = peak_prominences(jerk, peaks[, 2]),
    distance = sqrt((max(height) - height)^2 + (max(prominence) - prominence)^2)
  ) %>% 
  arrange(dt)
dist_density <- density(jerk_peaks$distance)
dist_peaks <- pracma::findpeaks(dist_density$y)
interpeaks <- dist_peaks[1, 2]:dist_peaks[2, 2]
dist_valley <- pracma::findpeaks(-dist_density$y[interpeaks])
dist_thr <- dist_density$x[dist_valley[1, 2] + dist_peaks[1, 2] - 1]
jerk_peaks <- jerk_peaks %>%  
  mutate(distance_cat = cut(distance, 
                            c(-Inf, 
                              min(distance[distance > 0]) / 2, 
                              dist_thr, 
                              Inf),
                            labels = c("max",
                                       "keep",
                                       "reject")))
shape_scale <- scale_shape_manual(values = c(max = 6, keep = 16, reject = 1))
hp_labs <- labs(x = "Peak height", y = "Peak prominence")

# Using region 14 as an example
jerk_example <- filter(bw_400hz, region_id == 14)
peak_example <- jerk_peaks %>% 
  filter(between(dt, first(jerk_example$dt), last(jerk_example$dt))) %>% 
  mutate(i = row_number())

# Labeled peaks in jerk
pA <- ggplot(jerk_example, aes(dt, njerk_smooth)) +
  geom_line() +
  geom_point(aes(shape = distance_cat), peak_example) +
  ggrepel::geom_text_repel(aes(label = i), peak_example) +
  shape_scale +
  expand_limits(y = 0) +
  labs(y = bquote(Jerk~(m~s^{-3}))) +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
# Peak height, prominence
pB <- ggplot(peak_example, aes(height, prominence, shape = distance_cat)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = i), max.overlaps = Inf) +
  shape_scale +
  hp_labs +
  coord_fixed() +
  theme_classic() +
  theme(legend.position = "none")

# All heights, prominences (with threshold)
x0 <- max(jerk_peaks$height)
y0 <- max(jerk_peaks$prominence)
r <- dist_thr
theta <- seq(0, 2 * pi, length.out = 256)
threshold <- tibble(
  height = x0 + r * cos(theta),
  prominence = y0 + r * sin(theta)
) %>% 
  filter(height <= x0,
         prominence <= y0)

pC <- jerk_peaks %>% 
    mutate() %>% 
  ggplot(aes(height, prominence)) +
  geom_point(aes(shape = distance_cat, size = distance_cat)) +
  annotate("path", 
           x = threshold$height, y = threshold$prominence,
           color = "red", linetype = "dashed") +
  shape_scale +
  scale_size_manual(values = c(max = 3, keep = 1, reject = 1)) +
  hp_labs +
  coord_fixed() +
  theme_classic() +
  theme(legend.position = "none")

pD <- ggplot(jerk_peaks, aes(distance)) +
  geom_histogram(aes(y = ..density.., fill = distance < dist_thr), 
                 bins = 25, color = "black") +
  geom_density(color = "blue") +
  geom_vline(xintercept = dist_thr, color = "red", linetype = 2) +
  scale_fill_manual(values = c(`TRUE` = "black", `FALSE` = "white")) +
  labs(x = "Distance from max") +
  theme_classic() +
  theme(legend.position = "none")

cowplot::plot_grid(pA, pB, pC, pD,
                   align = "h",
                   axis = "bt",
                   nrow = 2,
                   labels = "AUTO")
```
